Handbook streamFind: backend

### 2022-08-11 handbooks/Handbook_backend_streamFind.Rmd
LFU14

The streamFind package includes example files, which are located  in the folder `system.file(package = "streamFind", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, _m/z_ for MS1 data between 200 and 450 Da and _m/z_ for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, _m/z_ for MS1 data between 200 and 300 Da and _m/z_ for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the backend.

```{r resource-files, cache=FALSE}
#path to files
r_path <- system.file(package = "streamFind", dir = "extdata")

#list of file full paths
files <- list.files(r_path, pattern = ".mzML", full.names = TRUE)
```

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  N. = seq_len(length(files)),
  file_type = rep("MS", 6 + 18),
  file_name = basename(files),
  device = c(rep("Agilent Q-TOF", 24)),
  data_type = c(rep("centroid", 3), rep("profile", 3), rep("centroid", 18)),
  ac_mode = c(rep("MS/MS", 24)),
  polarity = c(rep("positive", 6),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3)),
  description = c(
    rep("Basic centroided MS data spiked with chemical and internal standards (CS and IS, respectively).", 3),
    rep("Basic profile MS data spiked with CS and IS.", 3),
    rep("Blank MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent treated with ozonated strong water MS data spiked with IS.", 6)
  )
)

@@

```{r create-msanalysis, message=FALSE, results='hide'}
#creating an msAnalysis object for the 4th file, corresponding to profile MS data
aProf <- newAnalysis(file = files[4])
```

@@

As the method `getAnalysis` can be used to obtain an _msAnalysis_ from an _msData_ object, the method `addAnalyses` is used to add extra _msAnalysis_ objects to _msData_. Below the files 4 to 6 are added to the already created _msData_ (containing files 1 to 3). Note that both `getAnalyses` and `addAnalyses` can be applied to equivalent classes to _msData_ (i.e., _uvData_ and _ramanData_). Also, note that when adding analyses to an _msData_ object with features (i.e., with peaks grouped across analyses), the _msFeatures_ content is cleared as the grouping is invalidated with new _msAnalysis_ objects.

@@

#all profile MS files
aProfs <- lapply(files[4:6], function(x) newAnalysis(x))

@@

* metadata of _msAnalysis_:

```{r metadata_structure}
#metadata of the fourth analyses in the msData
t(metadata(sp1, analyses = 4))

#applied directly to the msAnalysis S4 class
t(metadata(a1))
```

@@

add metadata to an _msAnalysis_ object (a1):

@@

metadata(a1, which = names(meta_vec))

@@

metadata(a1, which = names(meta_vec))
```

add metadata to an _msData_ object (sp1):

@@

metadata(sp1, which = names(meta_1))

@@

metadata(sp1, which = colnames(meta_df_2))
```

@@

@@

```{r newstreamset-basicworkflow, message=FALSE, results='hide'}
sp2 <- newStreamSet(files = files[c(10:12, 16:18)], path = getwd(), title = "Project Example")
```

@@

```{r other-s4methods}
#getter for analysis names in msdata
analysisNames(sp2)

@@

#getter for polarity of analyses in a msData
polarities(sp2)

>>>>

The streamFind package includes example data files via the streamFindData R package. Once installed the full file paths of MS data files can be obtained as shown below. Alternatively, the files can be obtained directly from folder `system.file(package = "streamFindData", dir = "extdata")`. The files available are described in Table \@ref(tab:description-files). Note, the MS files (.mzML) were trimmed with the function `trimSpectraFilesMZR` to reduce their size. The centroided Agilent Q-TOF files were trimmed with retention time between 900 and 1350 seconds, _m/z_ for MS1 data between 200 and 450 Da and _m/z_ for MS2 between 35 and 450 Da. Additionally, MS1 and MS2 traces below 100 and 50 counts were removed, respectively. The profile Agilent Q-TOF files were trimmed with retention time between 1000 and 1200 seconds, _m/z_ for MS1 data between 200 and 300 Da and _m/z_ for MS2 between 35 and 300 Da. Additionally, MS1 and MS2 traces below 1 count were removed to exclude empty spectra. The files are used within the document for demonstration of the backend.

```{r resource-files, cache=FALSE}

r_path <- system.file(package = "streamFindData", dir = "extdata")

files <- list.files(r_path, pattern = ".mzML|.mzXML", full.names = TRUE)

#list of file full paths from the data package
#files <- streamFindData::msFilePaths()
```

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  N. = seq_len(length(files)),
  file_type = rep("MS", 27),
  file_name = basename(files),
  device = c(rep("Agilent Q-TOF", 27)),
  data_type = c(rep("centroid", 6), rep("profile", 3), rep("centroid", 18)),
  ac_mode = c(rep("MS/MS", 27)),
  polarity = c(rep("positive", 9),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3)),
  description = c(
    rep("Basic centroided MS data as mzML spiked with chemical 
        and internal standards (CS and IS, respectively).", 3),
    rep("Basic centroided MS data as mzXML spiked with chemical 
        and internal standards (CS and IS, respectively).", 3),
    rep("Basic profile MS data as mzML spiked with CS and IS.", 3),
    rep("Blank MS data as mzML spiked with IS.", 6),
    rep("Wastewater secondary effluent MS data as mzML spiked with IS.", 6),
    rep("Wastewater secondary effluent treated with ozonated 
        strong water MS data as mzML spiked with IS.", 6)
  )
)

@@

```{r create-msanalysis, message=FALSE, results='hide'}
#creating an msAnalysis object for the 4th file, corresponding to profile MS data
aProf <- newAnalysis(file = files[7])
```

@@

As the method `getAnalysis` can be used to obtain an _msAnalysis_ from an _msData_ object, the method `addAnalyses` is used to add extra _msAnalysis_ objects to _msData_. Below the files 7 to 9 are added to the already created _msData_ (containing files 1 to 3). Note that both `getAnalyses` and `addAnalyses` can be applied to equivalent classes to _msData_ (i.e., _uvData_ and _ramanData_). Also, note that when adding analyses to an _msData_ object with features (i.e., with peaks grouped across analyses), the _msFeatures_ content is cleared as the grouping is invalidated with new _msAnalysis_ objects.

@@

#all profile MS files
aProfs <- lapply(files[7:9], function(x) newAnalysis(x))

@@

* get metadata:

```{r metadata_structure}
#metadata of the fourth analyses in the msData
getMetadata(sp1, analyses = 4)

#applied directly to the msAnalysis S4 class
getMetadata(a1)
```

@@

* add metadata to an _msAnalysis_ object (a1):

@@

getMetadata(a1, which = names(meta_vec))

@@

getMetadata(a1, which = names(meta_vec))
```

* add metadata to an _msData_ object (sp1):

@@

getMetadata(sp1, which = names(meta_1))

@@

getMetadata(sp1, analyses = c(3, 4), which = names(meta_1))
```

@@

`r # TODO add faster functionality for TIC and BPC based on xml2``

@@

```{r newstreamset-basicworkflow, message=FALSE, results='hide'}
sp2 <- newStreamSet(files = files[c(13:15, 19:21)], path = getwd(), title = "Project Example")
```

@@

```{r other-s4methods}
#getter for the analysis table
analysisTable(sp2)

#getter for analysis names in msdata
analysisNames(sp2)

@@

#getter for polarity of analyses in a msData
#polarities(sp2)
